---
title: "michelle"
format: html
editor: visual
execute: 
  warning: false
  message: false
html:
    embed-resources: true
    toc: true
---

## Background

### Methods according to Rennie et al. 2019

BACI: Before-after control-impact (BACI) design ANOVA:

-   Years were classified as either before aquaculture (prior to 2003)

-   During aquaculture (2003–2007)

-   After aquaculture (2008 and later).

-   A significant interaction term indicated difference in direction and/or magnitude of effects between the lake undergoing aquaculture and reference lake

-   Minnow counts (fathead minnow and all species combined) were summed daily and a catch-per-unit-effort (CUE) from trap nets (number of fish per net day, or number of fish divided by the product of the number of nets set and days that nets fished) was estimated for each spring and autumn sampling period.

To evaluate the impact of aquaculture on minnow abundance (mean CUE in either spring or autumn), the same BACI ANOVA was used as described above.

Minnow abundance was squareroot transformed to normalise and homogenise residual distributions. Pre-impact minnow abundance from 1989–2002 to provide sufficient sample sizes.

Separate ANOVAs were conducted in each season (spring and autumn).

Visual plots and Tukey post hoc tests were used to determine the significance and directionality of reported differences

### Results

Minnow abundance (mean seasonal CUE) in autumn increased sharply in Lake 375 during aquaculture, compared to lake 373 (ANOVA, lake by period interaction, F2,32 = 4.8, p = 0.015).

During aquaculture, autumn minnow abundance in Lake 375 was significantly higher than before (Tukey, p = 0.0004) and after aquaculture (Tukey, p = 0.04).

Minnow abundance in Lake 375 was significantly greater than in Lake 373 during aquaculture (Tukey, p = 0.005), but not different from 373 before and after aquaculture (Tukey, p ≫ 0.05). By contrast, there were no significant differences in minnow abundance between aquaculture and non-aquaculture years among lakes during spring (ANOVA, F2,32 = 0.3, p = 0.7; Figure 4b), suggestive of significant overwinter mortality of minnows. Increases in minnow abundance in Lake 375 were almost entirely driven by an increase in fathead minnow (Supporting Information Appendix B)

### Interpreting Results

What is an F-value?

-   Ratio of variability

-   F = Variance between / Variance within

-   Higher F-value = stronger differences between groups

-   Notation: (df between, df within)

    -   Between: DF of factor being tested (3 periods - 1)

    -   Within: DF for residuals (sample size - \# groups or df.residual on fall_model)

What is Turkey? - Tukey

-   Turkey's Honest Significant Difference (HSD) is a post-hoc test

-   Identifies which group means differ significantly after running ANOVA

-   Controls for type 1 error (false positives)

-   Printed results

    -   diff = difference in means

    -   lwr, upr, = lower and upper bounds for 95% CI

    -   p adj = adjusted p-value

## Setup

#### Load libraries

```{r}
library(here)
library(tidyverse)
library(janitor)

library(car) # For anova
```

#### Read in and clean data

```{r}
# Note that I renamed the data file
minnow <- read_csv(here("data", "minnow.csv")) %>%
  clean_names() %>%
  # Rename columns to fish species
  rename(
    northern_redbelly_dace = f182,
    finescale_dace = f183,
    fathead_minnow = f209,
    northern_pearl_dace = f214,
    slimy_sculpin = f382
  ) 
# Note: Possibly drop comments column? 
# Seasons also look weird: "aSpring" "zFall" 
# Location refers to net type? "Deep Net"    "Shallow Net" "C. L. Net"   "C.L. Net"    "Shallow"     "Deep"        "Wing net" 
# Flag is all NA
```

```{r}
# BACI attempt

# Assign year to time periods
minnow_baci <- minnow %>%
    mutate(
    period = case_when(
      year >= 1998 & year <= 2002 ~ "pre",
      year >= 2003 & year <= 2007 ~ "treatment",
      year >= 2008 & year <= 2013 ~ "post",
      TRUE ~ NA_character_  # Make NA outside the range
    )) %>%
# Sqrt minnow abundances
  mutate(sqrt_minnow = sqrt(minnows),
         sqrt_fathead_minnow = sqrt(fathead_minnow)) %>%
  mutate(lake = as.factor(lake))

  
# Separate data into seasons
fall <- minnow_baci %>%
  filter(season == 'zFall')

spring <- minnow_baci %>%
  filter(season == 'aSpring')
```

```{r}
# Test out BACI for fall

# Fit the model
fall_model <- aov(sqrt_minnow ~ lake * period, data = fall)
fall_anova <- Anova(fall_model, type = "III") 

# Print results
print(fall_model)
print(fall_anova)

# Check reference
#print(levels(fall$lake)) # 373 is reference
```

### Interpreting Results

SS: Variability in sqrt_minnow due to these components

RSS: Unexplained variation – large

RSE: Spread of residuals

Anova Table:

SS: Variation explained

F-value: variance explained by interaction / unexplained var (residuals)

(bigger = more explanation by the interaction)

Lake: p-value is 0.74 (\>\> 0.05) there is no sig. diff between lakes

Period: p-value is 0.34 (\>\>0.05) time period does not impact

Lakexperiod: p-value = 0.0001984; period on abundance differs between lakes

```{r}
# Turkey HSD for fall
turkey_fall <- TukeyHSD(fall_model, "lake:period")

# Print results
print(turkey_fall)

# Visualize results
plot(turkey_fall)

```

Significant values:

-   375:treatment vs. 373:post

    -   375 has more abundance than 373 after treatment

-   375:treatment vs. 375:post:

    -   375 has more abundance during treatment than after

-   375:treatment vs. 373:treatment

    -   375 during treatment has more abundance than 373 during treatment
